---
title: "Interpretability for Tree Ensembles"
subtitle: "Biostat 212A"
author: "Dr. Jin Zhou @ UCLA"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    cache: false

---

Credit: This note builds on tree/ensemble material from [_An Introduction to Statistical Learning: with Applications in R_](https://www.statlearning.com/) (ISL2) and [_Elements of Statistical Learning_](https://hastie.su.domains/ElemStatLearn/) (ESL2), and introduces modern interpretability tools including **permutation importance** and **SHAP**.

Display system information for reproducibility.

::: {.panel-tabset}
## R

```{r}
sessionInfo()
```

## Python

```{python}
import IPython
print(IPython.sys_info())
```
:::

# The problem: ensembles are black boxes

A single decision tree is readable --- you can trace the splitting rules and explain any prediction in a few steps.

But the models that *win* in practice are ensembles (random forests, boosting), which combine hundreds of trees. A single tree says: "If `CRuns > 300` and `CRBI > 200`, predict high salary." A 500-tree random forest just says: "The predicted salary is \$842K" --- *but why?*

We need tools to look inside. Two questions drive everything that follows:

1. **Which variables matter?** (variable importance)
2. **How does each variable affect a specific prediction?** (SHAP)

::: {.callout-note appearance="minimal"}
### Importance $\neq$ causality

A variable can be "important" because it correlates with the outcome, correlates with a true cause, or even due to data leakage. Importance answers: "Does the model rely on this variable?" --- not "Does this variable cause the outcome?"
:::


# Variable importance: two flavors

## Split / impurity importance (built into trees)

Every time a tree splits on feature $j$, the impurity (RSS, Gini, entropy) decreases by some amount $\Delta I(t)$. Sum those decreases over all splits on feature $j$, across all trees:

$$
VI_j = \sum_{t:\, \text{split variable}=j} \Delta I(t).
$$

This is computed for free during training and available in every tree library. But it can be **biased** toward high-cardinality features and tends to **split credit** unpredictably among correlated predictors.

## Permutation importance (model-agnostic)

A cleaner idea: measure how much the model's performance *drops* when you destroy a feature's information.

1. Score the model on held-out data → baseline metric $m_0$.
2. Randomly shuffle feature $j$ (break its link with $y$).
3. Re-score → $m_j$.
4. Importance $= m_0 - m_j$.

This works for any model and any metric. The main caveat: with correlated features, shuffling $X_j$ may not hurt much because $X_k$ substitutes for it.

::: {.callout-warning appearance="minimal"}
### The correlated-features trap

If $X_1$ and $X_2$ carry nearly the same information, **split importance** shares credit roughly 50/50 (neither looks dominant), while **permutation importance** underestimates both (each substitutes for the other). No importance method fully resolves this ambiguity --- keep a correlation matrix handy.
:::

## Worked example: `Hitters` salary data

::: {.panel-tabset}

#### R (tidymodels + ranger)
```{r}
library(tidymodels)
library(vip)
library(ISLR2)

Hitters2 <- Hitters %>% drop_na(Salary)

set.seed(1)
hitters_split <- initial_split(Hitters2, prop = 0.7)
hitters_train <- training(hitters_split)
hitters_test  <- testing(hitters_split)

# Two RF fits: one tracks impurity importance, one tracks permutation importance
rf_impurity_spec <-
  rand_forest(trees = 1000) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

rf_perm_spec <-
  rand_forest(trees = 1000) %>%
  set_engine("ranger", importance = "permutation") %>%
  set_mode("regression")

rf_impurity_fit <- workflow() %>% add_model(rf_impurity_spec) %>%
  add_formula(Salary ~ .) %>% fit(data = hitters_train)
rf_perm_fit <- workflow() %>% add_model(rf_perm_spec) %>%
  add_formula(Salary ~ .) %>% fit(data = hitters_train)

vip(extract_fit_engine(rf_impurity_fit), num_features = 10) +
  ggtitle("Split/impurity importance")
vip(extract_fit_engine(rf_perm_fit), num_features = 10) +
  ggtitle("Permutation importance (OOB)")
```

#### Python (sklearn)
```{python}
import pandas as pd
import numpy as np
from sklearn.model_selection import train_test_split
from sklearn.compose import ColumnTransformer
from sklearn.pipeline import Pipeline
from sklearn.preprocessing import OneHotEncoder
from sklearn.impute import SimpleImputer
from sklearn.ensemble import RandomForestRegressor
from sklearn.inspection import permutation_importance

Hitters = pd.read_csv("../data/Hitters.csv")
Hitters2 = Hitters.dropna(subset=["Salary"]).copy()

y = Hitters2["Salary"].values
X = Hitters2.drop(columns=["Salary"])

cat_cols = X.select_dtypes(include=["object", "category"]).columns.tolist()
num_cols = [c for c in X.columns if c not in cat_cols]

try:
    ohe = OneHotEncoder(handle_unknown="ignore", sparse_output=False)
except TypeError:
    ohe = OneHotEncoder(handle_unknown="ignore", sparse=False)

preprocess = ColumnTransformer(
    transformers=[
        ("num", Pipeline([("impute", SimpleImputer(strategy="median"))]), num_cols),
        ("cat", Pipeline([
            ("impute", SimpleImputer(strategy="most_frequent")),
            ("onehot", ohe)
        ]), cat_cols),
    ],
    remainder="drop",
    verbose_feature_names_out=False
)

rf = RandomForestRegressor(n_estimators=1000, random_state=1, n_jobs=-1)
pipe = Pipeline(steps=[("prep", preprocess), ("model", rf)])

X_train, X_test, y_train, y_test = train_test_split(
    X, y, test_size=0.3, random_state=1
)
pipe.fit(X_train, y_train)

# Permutation importance on held-out data
perm = permutation_importance(pipe, X_test, y_test, n_repeats=10,
                               random_state=1, n_jobs=-1)

feat_names = pipe.named_steps["prep"].get_feature_names_out()
top_idx = np.argsort(perm.importances_mean)[::-1][:10]
list(zip(feat_names[top_idx], np.round(perm.importances_mean[top_idx], 2)))
```
:::


# SHAP: local explanations that add up

## The idea 

Permutation importance gives one number per feature (global). **SHAP** gives one number per feature **per observation** (local). For a single prediction:

$$
\hat f(x) = \underbrace{\phi_0}_{\text{baseline}} + \phi_1 + \phi_2 + \cdots + \phi_p.
$$

Each $\phi_j$ says: "Feature $j$ pushed this prediction up (or down) by this much."

SHAP values come from **Shapley values** in game theory --- a principled way to fairly divide credit among players in a cooperative game.

**Group project analogy.** Three students (Alice, Bob, Carla) earn a group grade. Different subsets would have earned different grades:

| Coalition | Grade |
|---|---|
| nobody | 50 |
| {Alice} | 70 |
| {Bob} | 60 |
| {Alice, Bob} | 85 |
| {Alice, Bob, Carla} | 92 |
| ... | ... |

To fairly credit Alice: look at every possible ordering the students could have joined; record Alice's marginal contribution (grade jump when she joins); average across all orderings. That average is her Shapley value.

In ML: players = features, grade = model prediction. For tree ensembles, an algorithm called **TreeSHAP** computes this efficiently without enumerating all $2^p$ subsets.

## Key SHAP plots

- **Beeswarm**: each dot = one observation; $x$-axis = SHAP value; color = feature value. Shows which features matter *and* the direction of their effect.
- **Waterfall**: one observation explained step-by-step from baseline to final prediction.

## Worked example (boosted trees → SHAP)

::: {.panel-tabset}

#### R (xgboost + shapviz)
```{r}
library(tidymodels)
library(shapviz)

set.seed(1)

hitters_xgb_rec <-
  recipe(Salary ~ ., data = hitters_train) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

hitters_xgb_prep <- prep(hitters_xgb_rec)
p <- juice(hitters_xgb_prep) %>% select(-Salary) %>% ncol()

xgb_spec <-
  boost_tree(trees = 500, tree_depth = 3, learn_rate = 0.05,
             sample_size = 0.8, mtry = max(1, floor(0.8 * p))) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

xgb_fit <- workflow() %>% add_recipe(hitters_xgb_rec) %>%
  add_model(xgb_spec) %>% fit(data = hitters_train)

xgb_engine <- extract_fit_engine(xgb_fit)
X_train_mat <- bake(hitters_xgb_prep, new_data = hitters_train) %>%
  select(-Salary) %>% as.matrix()

sv <- shapviz(xgb_engine, X_pred = X_train_mat)

# Global: which features matter most?
sv_importance(sv, kind = "beeswarm")

# Local: explain one player's predicted salary
sv_waterfall(sv, row_id = 1)
```

#### Python (sklearn + shap)
```{python}
try:
    import shap
except ImportError:
    import sys, subprocess
    subprocess.check_call([sys.executable, "-m", "pip", "install", "shap"])
    import shap

from sklearn.ensemble import GradientBoostingRegressor

gbr = GradientBoostingRegressor(random_state=1)
pipe_gbr = Pipeline(steps=[("preprocess", preprocess), ("model", gbr)])
pipe_gbr.fit(X_train, y_train)

# Extract transformed features for SHAP
X_train_trans = pipe_gbr.named_steps["preprocess"].transform(X_train)
X_test_trans  = pipe_gbr.named_steps["preprocess"].transform(X_test)
feat_names = pipe_gbr.named_steps["preprocess"].get_feature_names_out()

X_train_shap = pd.DataFrame(X_train_trans, columns=feat_names)
X_test_shap  = pd.DataFrame(X_test_trans,  columns=feat_names)

model = pipe_gbr.named_steps["model"]

# TreeSHAP: background = training data, explain = test observations
explainer = shap.Explainer(model, X_train_shap)
shap_values = explainer(X_test_shap)

# Global: beeswarm
shap.plots.beeswarm(shap_values, max_display=10)

# Local: waterfall for one test observation
shap.plots.waterfall(shap_values[0])
```
:::


# Quick reference

| | Split / Gain | Permutation | SHAP |
|---|---|---|---|
| **What it gives** | Importance ranking | Importance ranking | Local + global explanations |
| **Global or local?** | Global | Global | Both |
| **Model-agnostic?** | Trees only | Any model | Practical for trees; possible for others |
| **Correlation caveat** | Shares credit | Underestimates (substitution) | Shares credit |
| **Speed** | Free | Moderate | Fast for trees (TreeSHAP) |

### Recommended workflow

1. Fit your tree ensemble (RF or boosting).
2. **Split importance** for a quick first look; **permutation importance** for a more reliable ranking.
3. **SHAP beeswarm** to see direction of effects; **waterfall** to explain individual predictions.
